---
title: "The Factors Affecting University Student Grades"
author: 'Group No: 12'
date: "2024-06-23"
output:
  html_document: default
  pdf_document: default
---

# Data Exploration


#### Load the Libraries

```{r }

# Load the necessary libraries
library(dplyr)
library(caret)
library(ggplot2)
library(corrplot)
library(reshape2) 

```

#### Load the Dataset

```{r }

# Load the data set
dataset <- read.csv(file.choose(), header = TRUE)

```

#### Structure of the Dataset

```{r }

# Display structure of the data set
str(dataset)

```

#### Summary Statistics

```{r }

# Summary statistics of the data set
summary(dataset)

```

#### Handling Missing Values

```{r }

# Identify numerical variables in the dataset
numerical_var <- sapply(dataset, is.numeric)

# Extract numerical columns
numerical_cols <- dataset[, numerical_var]

# Count missing values in each numerical column
colSums(is.na(numerical_cols))

# Replace missing values with each column median
numerical_cols <- numerical_cols %>%
  mutate_all(~ ifelse(is.na(.), median(., na.rm = TRUE), .))

# Verify missing values are handled 
colSums(is.na(numerical_cols))

# Replace original numerical columns in the dataset with imputed values
dataset[, names(numerical_cols)] <- numerical_cols


# Identify categorical variables in the dataset
categorical_var <- sapply(dataset, is.factor) | sapply(dataset, is.character)

# Extract categorical columns
categorical_cols <- dataset[, categorical_var]

# Count empty strings in each categorical column
colSums(categorical_cols == "")

# Define the impute_mode function to find the mode
impute_mode <- function(x) {
  unique_values <- unique(x)
  mode_value <- unique_values[which.max(tabulate(match(x, unique_values)))]
  return(mode_value)
}

# Replace empty strings with impute_mode function
categorical_cols <- categorical_cols %>%
  mutate_all(~ ifelse(. == "", impute_mode(.), .))

# Verify empty strings are handled
colSums(categorical_cols == "")

# Replace original categorical columns in the dataset with imputed values
dataset[, names(categorical_cols)] <- categorical_cols

```

# Data Preprocessing


#### Handling Outliers

```{r }

# Define a function to identify outliers using IQR method
numerical_outlier_detection  <- function(x) {
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  return(x[x < lower_bound | x > upper_bound])
}

# Apply numerical_outlier_detection function to each numerical column
numerical_outliers <- lapply(numerical_cols, numerical_outlier_detection)

# Print the numerical Outliers
numerical_outliers
```

### Objective 1: Evaluate the impact of demographic factors such as Age, Gender, Financial Status on the GPA of university students.

```{r }

# Summary statistics for GPA by Age
age_gpa_summary <- dataset %>%
  group_by(Age) %>%
  summarize(
    Mean_GPA = mean(GPA, na.rm = TRUE),
    SD_GPA = sd(GPA, na.rm = TRUE),
    Min_GPA = min(GPA, na.rm = TRUE),
    Max_GPA = max(GPA, na.rm = TRUE)
  )
print(age_gpa_summary)


# Summary statistics for GPA by Gender
gender_gpa_summary <- dataset %>%
  group_by(Gender) %>%
  summarize(
    Mean_GPA = mean(GPA, na.rm = TRUE),
    SD_GPA = sd(GPA, na.rm = TRUE),
    Min_GPA = min(GPA, na.rm = TRUE),
    Max_GPA = max(GPA, na.rm = TRUE)
  )
print(gender_gpa_summary)


# Summary statistics for GPA by Financial Status
financial_gpa_summary <- dataset %>%
  group_by(Financial_Status) %>%
  summarize(
    Mean_GPA = mean(GPA, na.rm = TRUE),
    SD_GPA = sd(GPA, na.rm = TRUE),
    Min_GPA = min(GPA, na.rm = TRUE),
    Max_GPA = max(GPA, na.rm = TRUE)
  )
print(financial_gpa_summary)

```

### Objective 2: Examine the relationship between psychological factors such as Motivation, Bullying, Stress Levels, Lack of Interest, learning Style on the GPA of university students.

```{r }

# Visualize the relationship between Bullying and GPA
ggplot(dataset, aes(x = Bullying, y = GPA)) +
  geom_boxplot() +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = round(after_stat(y), 2)),
    vjust = -1.5,
    color = "blue"
  ) +
  stat_summary(
    fun.data = function(y) {
      data.frame(
        y = quantile(y, 0.75),
        label = paste("IQR:", round(IQR(y), 2))
      )
    },
    geom = "text",
    vjust = -0.5,
    color = "red"
  ) +
  labs(title = "GPA by Bullying", x = "Bullying", y = "GPA")


# Visualize the relationship between Motivation and GPA
ggplot(dataset, aes(x = Motivation, y = GPA)) +
  geom_boxplot() +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = round(after_stat(y), 2)),
    vjust = -1.5,
    color = "blue"
  ) +
  stat_summary(
    fun.data = function(y) {
      data.frame(
        y = quantile(y, 0.75),
        label = paste("IQR:", round(IQR(y), 2))
      )
    },
    geom = "text",
    vjust = -0.5,
    color = "red"
  ) +
  labs(title = "GPA by Motivation", x = "Motivation", y = "GPA")


# Visualize the relationship between Stress_Levels and GPA
ggplot(dataset, aes(x = Stress_Levels, y = GPA)) +
  geom_boxplot() +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = round(after_stat(y), 2)),
    vjust = -1.5,
    color = "blue"
  ) +
  stat_summary(
    fun.data = function(y) {
      data.frame(
        y = quantile(y, 0.75),
        label = paste("IQR:", round(IQR(y), 2))
      )
    },
    geom = "text",
    vjust = -0.5,
    color = "red"
  ) +
  labs(title = "GPA by Stress Levels", x = "Stress Levels", y = "GPA")

# Visualize the relationship between Lack_of_Interest and GPA
ggplot(dataset, aes(x = Lack_of_Interest, y = GPA)) +
  geom_boxplot() +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = round(after_stat(y), 2)),
    vjust = -1.5,
    color = "blue"
  ) +
  stat_summary(
    fun.data = function(y) {
      data.frame(
        y = quantile(y, 0.75),
        label = paste("IQR:", round(IQR(y), 2))
      )
    },
    geom = "text",
    vjust = -0.5,
    color = "red"
  ) +
  labs(title = "GPA by Lack of Interest", x = "Lack of Interest", y = "GPA")


# Visualize the relationship between Learning_Style and GPA
ggplot(dataset, aes(x = Learning_Style, y = GPA)) +
  geom_boxplot() +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = round(after_stat(y), 2)),
    vjust = -1.5,
    color = "blue"
  ) +
  stat_summary(
    fun.data = function(y) {
      data.frame(
        y = quantile(y, 0.75),
        label = paste("IQR:", round(IQR(y), 2))
      )
    },
    geom = "text",
    vjust = -0.5,
    color = "red"
  ) +
  labs(title = "GPA by Learning Style", x = "Lack of Learning Style", y = "GPA")


```

### Objective 3: Analyze the correlation between lifestyle factors such as Sleep Hours, Screen Time on the GPA of university students.

```{r }

# Correlation between Sleep_Hours_PerDay and GPA 
cor(dataset$Sleep_Hours_PerDay, dataset$GPA, use = "complete.obs")

# Correlation between Screen_Time_PerDay and GPA 
cor(dataset$Screen_Time_PerDay, dataset$GPA, use = "complete.obs")

```

### Objective 4: The influence of academic-related factors such as Previous GPA, Attendance, Study Hours, Class Size on the GPA of university students.

```{r }

# Standardize numerical variables
dataset[numerical_var] <- scale(dataset[numerical_var])

# Print the summary statistics of standardized numerical variables
summary(dataset[numerical_var])

```

```{r }

# Set seed for reproducibility
set.seed(123)

# Split the dataset into training and testing sets (80% train, 20% test)
train_index <- createDataPartition(dataset$GPA, p = 0.8, list = FALSE)
train_data <- dataset[train_index, ]
test_data <- dataset[-train_index, ]

# Fit multiple linear regression model
lm_model <- lm(GPA ~ Previous_GPA + Attendance + Study_Hours_PerWeek + Class_Size, data = train_data)

# Summarize the model
summary(lm_model)

```

```{r }

# Predict GPA using the test_data
predicted_gpa <- predict(lm_model, newdata = test_data)


# Calculate Mean Squared Error (MSE)
mse <- mean((test_data$GPA - predicted_gpa)^2)
cat("Mean Squared Error (MSE):", mse, "\n")


# Calculate Root Mean Squared Error (RMSE)
rmse <- sqrt(mse)
cat("Root Mean Squared Error (RMSE):", rmse, "\n")


# Extract R-squared value from the summary of the linear regression model
r_squared <- summary(lm_model)$r.squared
cat("R-Squared:", r_squared, "\n")

```

```{r }
# Create a data frame to compare actual and predicted values
comparison <- data.frame(Actual = test_data$GPA, Predicted = predicted_gpa)

# Visualize the comparison of actual vs predicted GPA
ggplot(comparison, aes(x = Actual, y = Predicted)) +
  geom_point() +
  labs(x = "Actual GPA", y = "Predicted GPA") +
  ggtitle("Actual vs Predicted GPA")

```